evidence of adaptation-independent time-warped spike timing noise correlations in cortical networks

### 2020-08-11 manuscript.tex
Smashing!

\usepackage[colorinlistoftodos]{todonotes}

% \usepackage[nomarkers,notablist,nofiglist]{endfloat}
% \usepackage[nomarkers,notablist,nofiglist]{endfloat}
% \usepackage[colorinlistoftodos,disable]{todonotes}

@@

\begin{abstract}
% http://blogs.nature.com/naturejobs/2014/11/03/how-to-get-published-in-high-impact-journals-big-research-and-better-writing/
% \color{blue}`Do not undersell your research. “If you find something big, tell us!'\color{Black}

REL: Efficient population coding of naturalistic whisker motion in the ventro-posterior medial thalamus based on precise spike timing

Sensory stimulation induced patterns of action potentials recorded in adult rat cortex in vivo allow investigation of neural coding principles.

Patterns of action potentials induced by sensory stimulation in vivo allow investigation of neural coding principles. Trial-to-trial variability engenders debate over temporal and rate codes. Characterising distributions of stimulus responses enables discovery of response properties, which reliably encode the stimulus and may affect other processes. Non-stationarity (i.e. trial dependence) and noise correlations complicate such characterisation. A novel algorithm found correlated clusters above chance in first spike stimulus-response distributions for pairs of barrel cortex units. First spike times covaried at angles different from $45^{\circ}$ for a proportion of stationary and non-stationary (after removal of trial-dependence) clusters, suggestive of an underlying factor warping the distribution of decodable relative spike time differences from trial-to-trial. Factor analysis allowed estimation of factor-dependent and trial-independent stimulus response distributions, with a high estimated potential for discriminable stimulus encoding. The results suggest that consideration of clustering, factor-dependence and non-stationarity, together, are necessary to explain a large proportion of neural variability.

\end{abstract}

\keywords{Neural coding, precise spike timing, time-warp, state-dependent processing, barrel cortex, noise correlations, neural adaptation, in vivo recording, feature-binding}

@@

It is possible that spike time variability is caused and modulated across a population by a low-dimensional underlying state. This study aims to characterise the nature of trial-to-trial spike time noise correlations as a step towards confirming or rejecting the modulation of spike times by an underlying low-dimensional latent state. This offers a novel prediction of the encoding capacity of state-dependent spike time encoding. These results control for the occurrence of neural adaptation. The demonstrated nature of spike time noise correlations demonstrated offers an opportunity to  explain the report of only weak noise correlations in the previous characterisations of this data set \cite{reyes2015high}. Characterising spike timing codes is important to understanding a variety of dynamic, homeostatic and plastic processes, as well as properties of action potential patterns directly involved in cognition.

@@

Under such approaches, `noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimulus \cite{averbeck2006neural}. `Signal correlations' concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}). Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}. For fixed-state decoders, noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}.

@@

Regardless of whether decoding neurons in the brain respond in a fixed-state or state-dependent manner, characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}. 
% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
Processes of adaptation, which we define as different to latent states underlying trial-to-trial noise correlations, could also lead to correlative structure in the stimulus-response distribution (Fig. X). Adaptation must therefore be controlled for when characterising adaptation-independent noise correlations.

@@

The nature of first spike patterns has been explored in previous research. First spike times have been observed with millisecond order precision relative to stimulus onset across species and modalities in sensory organs \cite{johansson2004first, uzzell2004precision, gollisch2008rapid}, thalamic areas \cite{reinagel2002precise, storchi2012comparison} and early sensory cortices \cite{reyes2014laminar}, as well as relative to oscillations \cite{havenith2011synchrony}. First spike times in the rat barrel cortex have been shown to improve stimulus encoding by 44\% over spike count \cite{panzeri2001role}. Precise relative differences between the spike times of multiple neurons \cite{izhikevich2006polychronization} do not require reference to an onset signal and have been observed experimentally \cite{johansson2004first, gollisch2008rapid, reyes2014laminar} and shown to encode stimulus information when averaged over trials \cite{gollisch2008rapid}.

@@

A previous study analysing the same data set found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}. Another study analysing the data set could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns but found only weak spike count noise correlations \cite{reyes2015high}. The complex nature of spike time correlations uncovered in this study for the same data set may therefore offer an improved opportunity to study population-wide modulation by low-dimensional latent states. The new analysis uses time series analysis to control for neural adaptation over all stimulus trials. Adaptation occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}, and was confirmed to occur for 7 out of 8 neuron groups in this study.

@@

For repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked (see plots in the top left of Supplementary Vid. 1). A novel algorithm extracted isolated clusters from these single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}. The novel algorithm, however, also showed that first spike pairs within these isolated clusters were positively correlated above chance-levels.

Moreover, the correlation angle of many of the correlated clusters was different from $45^{\circ}$. This cannot be explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle). Instead the `non-unitary' correlation gradients (different from $45^{\circ}$) may be indicative that correlations are due to an internal shared latent state. Clusters remained correlated with non-unitary gradient after controlling for neural adaptation using time series analysis, confirming the correlated clusters as evidence of precise non-adaptive noise correlations. A large proportion of these noise correlations would have been missed without clustering.

Importantly, the observation of adaptation-independent correlated clusters with non-unitary gradient shows that the distribution of decodable first spike time relative differences for pairs of neurons are `warped' from trial-to-trial by some `non-adaptive' continuous latent state. This phenomenon is referred to as `adaptation-independent clustered time-warped spike timing'. This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not consider discriminable encoding, clustering, adaptation and different warping functions for different neurons \cite{williams2020discovering}. One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons need not respond identically to warps of the same pattern. Rather, neurons should process relative spike patterns in a state-dependent manner to utilise the improved state-dependent coding capacity.

Adaptation-controlled non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers. This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner. Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaptation-controlled correlated cluster. This gave a prediction of the resolution of possible state-dependent encoding, which surpasses previously proposed coding forms.

The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed. Estimating whether cluster correlations observed for different pairs of neurons is modulated by a single or low-dimensional shared state was made difficult by variability between clusters in the trials from which cluster samples came [See notes below]. The work of this paper in extracting and characterising adaptation-independent correlated clusters provides a necessary component step towards confirming or rejecting state-dependent spike-time coding, and is also important under the classical fixed-state decoder interpretation. The results show that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.

@@

\subsection*{First spike pair response distributions}

@@

Analysing the joint responses of two or more units to repeated presentations of a single stimulus offers the opportunity to discern noise correlations caused by adaptative or non-adaptative processes. For the purpose of understanding, this section presents an example distribution of first spike pairs for a pair of units (L5AI and L4I) in the principle column in response to a 4Hz stimulation of the corresponding principle whisker (over a 50ms window following cortical activation onset). Only responses on `conjunctive trials' (trials in which both units spiked for the single stimulus) are shown.

First spike pairs appear to come from one of two clusters (blue and green dots) for presentations of the same stimulus. Moreover, the green cluster appears correlated with non-unitary gradient, suggestive that some underlying processes (adaptative or non-adaptative) may covary the first spike times of the two units. Such correlated clusters were observed in a large number of first spike pair distributions (Supplementary Vid. 1).

This paper aims to quantify the occurrence of such correlated clusters and determine when adaptative and/or non-adaptative processes underlie such noise correlations. Importantly, the apparent non-unitary gradient of correlation of the green cluster in Fig. 3a suggests that the distribution of differences varies with respect to the possible adaptative and/or non-adaptative processes underlying the correlation.

The orange Gaussian in Fig. 3a is centered at the mean of the first spike pairs and has widths equal to 4 times the first spike time standard deviations of each unit (on conjunctive trial), illustrating the region from which 99.9\% of first spike pairs would be predicted to fall under the assumption of independence and normality (i.e. spike time jitter). This assumption fails to capture the distribution structure of this and many other spike pair distributions (Supplementary Vid. 1). 
% Moreover, very few different stimuli could be encoded with high-discriminability.

% The spike time differences have a standard deviation of X (Fig. 3b), which is also poor for discriminable encoding.

Moreover, the apparent non-unitary gradient of correlation suggests that the distribution of differences varies with respect to the value of this underlying state, such that the state-conditioned distribution of relative first spike differences would be more compact than the general case, thus offering a higher potential capacity for stimulus encoding if downstream neurons could respond in a state-dependent manner.

% \subsection*{Clustering algorithm summary}

@@

A two-stage algorithm was developed to quantify the occurrence and angle of correlated clusters. Stage 1 of the algorithm extracted well isolated clusters defined by the $4\sigma$ boundary of flat 2D Gaussian ellipses (see Methods \ref{methods:flat_clusters}; e.g. flat solid green ellipse in Fig. 3a).
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary0}
The algorithm was verified by confirming that the algorithm extracted correlated clusters at chance levels for multiple control data sets (see Methods \ref{methods:correlation_surrogate_control_datasets}; Supplementary Fig. 3).

@@

>>>>

% \usepackage[colorinlistoftodos]{todonotes}

\usepackage[nomarkers,notablist,nofiglist]{endfloat}
\usepackage[colorinlistoftodos,disable]{todonotes}

@@

% \begin{abstract}
% % http://blogs.nature.com/naturejobs/2014/11/03/how-to-get-published-in-high-impact-journals-big-research-and-better-writing/
% % \color{blue}`Do not undersell your research. “If you find something big, tell us!'\color{Black}

% REL: Efficient population coding of naturalistic whisker motion in the ventro-posterior medial thalamus based on precise spike timing

% Sensory stimulation induced patterns of action potentials recorded in adult rat cortex in vivo allow investigation of neural coding principles.

% Patterns of action potentials induced by sensory stimulation in vivo allow investigation of neural coding principles. Trial-to-trial variability engenders debate over temporal and rate codes. Characterising distributions of stimulus responses enables discovery of response properties, which reliably encode the stimulus and may affect other processes. Non-stationarity (i.e. trial dependence) and noise correlations complicate such characterisation. A novel algorithm found correlated clusters above chance in first spike stimulus-response distributions for pairs of barrel cortex units. First spike times covaried at angles different from $45^{\circ}$ for a proportion of stationary and non-stationary (after removal of trial-dependence) clusters, suggestive of an underlying factor warping the distribution of decodable relative spike time differences from trial-to-trial. Factor analysis allowed estimation of factor-dependent and trial-independent stimulus response distributions, with a high estimated potential for discriminable stimulus encoding. The results suggest that consideration of clustering, factor-dependence and non-stationarity, together, are necessary to explain a large proportion of neural variability.

% \end{abstract}

% \keywords{Neural coding, precise spike timing, time-warp, state-dependent processing, barrel cortex, noise correlations, neural adaptation, in vivo recording, feature-binding}

@@

It is possible that spike time variability is caused and modulated across a population by such a low-dimensional continuous latent state. This study aims to characterise the nature of trial-to-trial spike time noise correlations as a step towards confirming or rejecting the modulation of spike times by a continuous low-dimensional latent state. This offers a novel prediction of the encoding capacity of state-dependent spike time encoding. These results control for neural adaptation. 
Characterising spike timing codes is important to understanding a variety of dynamic and plastic processes, including properties of action potential patterns directly involved in cognition.

@@

Under such approaches, `noise correlations' refer to shared trial-to-trial variability between the responses of multiple neurons over repeated presentations of single stimuli \cite{averbeck2006neural}. `Signal correlations' concern tuning similarity between neurons over multiple stimuli (see \cite{averbeck2006neural}). Neural coding and noise correlations are classically considered with respect to fixed-state decoders \cite{averbeck2006neural, moreno2014information, stringer2019high}.

@@

Characterising noise correlations within stimulus-response distributions is complicated by the heterogeneity of sample response distribution structure and neural adaptation, which occurs across species and modalities \cite{dragoi2002dynamics, ulanovsky2003processing, sharpee2006adaptive}. 
% That is, the response distributions are \textit{non-stationary} and change monotonically as a function of trial index for some neurons. 
Processes of adaptation, which we define as different to continuous latent states underlying trial-to-trial noise correlations, could also lead to correlative structure in the stimulus-response distribution (Fig. X). Adaptation must therefore be controlled for when characterising adaptation-independent noise correlations.

@@

First spike patterns have been explored in previous research. First spike times have been observed with millisecond order precision relative to stimulus onset across species and modalities in sensory organs \cite{johansson2004first, uzzell2004precision, gollisch2008rapid}, thalamic areas \cite{reinagel2002precise, storchi2012comparison} and early sensory cortices \cite{reyes2014laminar}, as well as relative to oscillations \cite{havenith2011synchrony}. First spike times in the rat barrel cortex have been shown to improve stimulus encoding by 44\% over spike count \cite{panzeri2001role}. Precise relative differences between the spike times of multiple neurons \cite{izhikevich2006polychronization} do not require reference to an onset signal and have been observed experimentally \cite{johansson2004first, gollisch2008rapid, reyes2014laminar} and shown to encode stimulus information when averaged over trials \cite{gollisch2008rapid}.

@@

A previous study analysing the same data set found precise multi-neuron spike patterns that were intra- and inter-laminar, as well intra- and intra-columnar \cite{reyes2014laminar}. Another study analysing the data set could decode stimulus properties (stimulation frequency and location) from spike counts and spike patterns but found only weak spike count noise correlations \cite{reyes2015high}. The spike time noise correlations uncovered in this study offer a new opportunity to study population-wide modulation by low-dimensional latent states and controls for neural adaptation, which occurs in the barrel cortex to repetitive whisker stimulation for both spike count \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response, khatri2009stimulus, martin2014tactile, barros2019response} and spike latency \cite{ahissar2000transformation, ahissar2001temporal, kheradpezhouh2017response}.
% and was confirmed to occur for 7 out of 8 neuron groups in this study.

@@

For repeated trials of a single stimulus, the first spike times of two units were plotted against each other for trials on which they both spiked. For example, Fig. X shows first spike pairs for a pair of units (L5AI and L4I) in the principle column in response to a 4Hz stimulation of the corresponding principle whisker. First spike pairs appear to come from one of two clusters (blue and green dots) for presentations of the same stimulus. Moreover, the green cluster appears correlated with non-unitary gradient (different from $45^{\circ}$).

Such correlated clusters were observed in a large number of first spike pair distributions (Supplementary Vid. 1) and represent reliable regions of the response distribution space that can be analysed to test the nature of spike timing noise correlations. The correlative structure could be caused by neural adaptation, a continuous non-adaptative latent state or a combination of the two.

% The paper aims to test that the occurrence of correlated clusters is above chance and to test the degree to which such correlations are caused by adaptative and non-adaptative processes.

% (Fig X; see plots in the top left of Supplementary Vid. 1).

A novel algorithm extracted isolated clusters from single stimulus first spike pair response distributions, demonstrating that temporally precise spike patterns fire on subsets of trials, as shown previously for the data set \cite{reyes2014laminar}. The algorithm, however, also showed that first spike pairs within these isolated clusters were positively correlated above chance-levels. A large proportion of these noise correlations would have been missed without clustering. Moreover, many of the correlated clusters had non-unitary correlation angles.

Importantly, clusters remained correlated with non-unitary gradient after controlling for neural adaptation,
% using time series analysis, 
giving evidence of precise non-adaptive noise correlations. Such non-unitary spike time correlation angles cannot be explained by a trial-to-trial shift in the onset of the stimulus or cortical activity, which would produce correlations with a unitary gradient (i.e. a $45^{\circ}$ angle). Instead the `non-unitary' correlation gradients may be indicative of covariation by a non-adaptative continuous latent state that varies from trial-to-trial.

This novel finding of `adaptation-independent clustered time-warped spike-timing noise correlations' shows that the distribution of decodable first spike time relative differences for pairs of neurons are `warped' from trial-to-trial by some `non-adaptive' continuous latent state.
% This phenomenon is referred to as `adaptation-independent clustered time-warped spike-timing noise correlations'. 
This differs from previous `time-warp' studies, which aim to align spike trains over trials and do not simultaneously consider stimulus encoding, clustering, adaptation and different warping functions for different neurons \cite{williams2020discovering}.

The characterisation of adaptation-independent clustered spike-timing noise correlations is useful under the fixed-state decoder interpretations where noise correlations can either increase or decrease encoding capacity \cite{averbeck2006neural}. More interestingly, these noise correlations open the possibility of state-dependent spike time coding. Factor analysis provided an estimate of state-dependent response distributions of first spike pairs (and relative first spike differences) for each adaptation-controlled correlated cluster. This gave a prediction of the resolution of possible state-dependent encoding, which surpasses previously proposed coding forms.

Adaptation-independent non-unitary correlated clusters appeared for pairs of neurons spatially separated over multiple barrels and cortical layers. This offers suggestive evidence of a shared low-dimensional continuous latent state affecting the spike times of spatially distanced neurons, which in turn could also modulate downstream neurons and allow them to respond in a state-dependent manner. 
% The number of low-dimensional continuous latent states underlying the correlative structure, however, must be confirmed. 
Estimating whether cluster correlations observed for different pairs of neurons are modulated by a single or low-dimensional shared state was made difficult by variability between clusters in the trials from which cluster samples came [See notes below]. The work of this paper in extracting and characterising adaptation-independent correlated clusters provides a necessary component step towards confirming or rejecting state-dependent spike-time coding. The results show that consideration of clustering, correlative structure and adaptation, together, are necessary to explain a large proportion of neural variability.

% Whether caused by adaptative and/or non-adaptative processes, the presence of correlated clusters with non-unitary correlation angles would be evidence that the distribution of relative spike time differences that representations of a stimulus come from are not fixed. If a continuous non-adaptative latent state underlies the correlations, then the state-dependent distribution of relative differences depends on the value of this state on a single trial. If neural adaptation underlies such correlations then the, than the the distribution of relative differences on a single trial covers a smaller region of space than the region from which responses come from over all trials. [HERE] If a combination of an adaptative and non-adaptative continuous latent state underlies the correlations then..

% The orange Gaussian in Fig. 3a is centered at the mean of the first spike pairs and has widths equal to 4 times the first spike time standard deviations of each unit (on conjunctive trial), illustrating the region from which 99.9\% of first spike pairs would be predicted to fall under the assumption of independence and normality (i.e. spike time jitter). This assumption fails to capture the distribution structure of this and many other spike pair distributions (Supplementary Vid. 1).

% Moreover, the apparent non-unitary gradient of correlation suggests that the distribution of differences varies with respect to the value of this underlying state, such that the state-conditioned distribution of relative first spike differences would be more compact than the general case, thus offering a higher potential capacity for stimulus encoding if downstream neurons could respond in a state-dependent manner.

@@

% \subsection*{First spike pair response distributions}

@@

@@

Analysing the joint responses of two or more units to repeated presentations of a single stimulus offers the opportunity to discern noise correlations caused by adaptative or non-adaptative processes.
A two-stage algorithm was developed to quantify the occurrence and angle of correlated clusters observed in first spike response distributions for pairs of units (Fig. X; top left of Supplementary Vid. 1). Stage 1 of the algorithm extracted well isolated clusters defined by the $4\sigma$ boundary of flat 2D Gaussian ellipses (see Methods \ref{methods:flat_clusters}; e.g. flat solid green ellipse in Fig. 3a).
\ExecuteMetaData[AnalysisOutputLatex.tex]{ClusterCorrelationSummary0}
The algorithm was verified by confirming that the algorithm extracted correlated clusters at chance levels for multiple control data sets (see Methods \ref{methods:correlation_surrogate_control_datasets}; Supplementary Fig. 3).

@@

\subsection*{Time warp integration}

One previous study has modelled time-warp-invariant integration \cite{gutig2009time}, although neurons need not respond identically to warps of the same pattern. Rather, neurons should process relative spike patterns in a state-dependent manner to utilise the improved state-dependent coding capacity.
